---
title: "586 Project - Oreoscoptes Analysis"
author: "ZiYing(Sophie) Chen, Xinyu Dong, Matthew Yau"
date: "2023-04-23"
output:
  html_document:
    css: tutorial.css
    fig_caption: yes
    highlight: textmate
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE,warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
# Load required packages
library(sp)
library(dplyr)
library(sf)
library(spatstat)
library(maptools)
library(rgdal)

# Load Oreoscoptes data
df <- read.csv("BC_data.csv", stringsAsFactors = FALSE)

# Generate a dataframe with longitude and latitude in BC_data
coord <- df[,c('decimalLatitude','decimalLongitude')]

# Clean the data by removing the observations with NA
coord <- na.omit(coord)

# Visualize the data 
# plot(coord$decimalLatitude, coord$decimalLongitude)

# Convert to CRS
coordinates(coord)<- ~decimalLongitude+decimalLatitude
    
proj4string(coord) <- CRS("+proj=longlat +datum=WGS84")
    
coord_conv <- spTransform(coord,CRS("+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000 +y_0=0 +datum=NAD83 +units=m +no_defs"))

# Load the covariates data
load("BC_Covariates.Rda")

# Define the observation area, window
win = as.owin(DATA$Window)

# create ppp object with coordination and the window 
ppp_birds <- ppp(coord_conv@coords[,1], 
                 coord_conv@coords[,2], 
                 window=win)

rho_elev <- rhohat(ppp_birds, DATA$Elevation)
rho_fore <- rhohat(ppp_birds, DATA$Forest)
rho_hfi <- rhohat(ppp_birds, DATA$HFI)
rho_dist_water <- rhohat(ppp_birds, DATA$Dist_Water)

# model building
mu_elev <- mean(DATA$Elevation)
stdev_elev <- sd(DATA$Elevation)
DATA$Elevation_scaled <- eval.im((Elevation - mu_elev)/stdev_elev, DATA)

mu_water <- mean(DATA$Dist_Water)
stdev_water <- sd(DATA$Dist_Water)
DATA$Dist_Water_scaled <- eval.im((Dist_Water - mu_water)/stdev_water, DATA)

# Fit the model with scaled covariates
formula1 <- ppp_birds ~ Elevation_scaled + I(Elevation_scaled^2) + Dist_Water_scaled + I(Dist_Water_scaled^2)
fit <- ppm(formula1, data = DATA)

formula2 <- ppp_birds ~ Elevation_scaled + Dist_Water_scaled + I(Dist_Water_scaled^2)
fit2 <- ppm(formula2, data = DATA)

co <- coef(fit2)

# Model Validation

#Run the quadrat test
quadrat.test(fit2, nx = 2, ny = 4)


```


## Introduction
The Sage Thrasher is a medium-sized passerine bird from the family Mimidae, which also includes mockingbirds, tremblers, and New World catbirds. It is the only member of the genus Oreoscoptes. This seems less close to the Caribbean thrashers, but rather to the mockingbirds instead.

The Sage Thrasher is one of the rarest birds in Canada. Prior to 1930, there were 30 or more nesting pairs at up to six sites; in recent years, only 4 - 10 pairs have been seen at one or two sites. Over the past 70 years, there has been an estimated 50 percent loss of habitat for the Sage Thrasher. Heavy grazing and clearing of sagebrush to improve range conditions has seriously affected the availability of suitable nesting areas. The preservation of remaining Sage Thrashers depends on the careful management of remaining large tracts of dense sagebrush habitat. Heavy grazing of sagebrush sites should be avoided, and areas cleared of sagebrush should be restored whenever possible.

The data used in this reports come from `eBirds`. `eBird` is a collective enterprise that takes a novel approach to citizen science by developing cooperative partnerships among experts in a wide range of fields: population ecologists, conservation biologists, quantitative ecologists, statisticians, computer scientists, GIS and informatics specialists, application developers, and data administrators.

Now we konw that `distance to water` and `elevation` plays a crucial role if we want to dig out where we could find these birds, which only mean this varible can help us even *predict* where this birds will live, based on theri current habitats.

This is of crucial meaning in identify potential location for re-habitation.  ....

## Methods  

### Data Description: Briefly describe the data and what variables are included. 


### Detailed Analytical Workflow

1. Load all packages we need. Please refer to **Statistical Packages**.

2. Data processing: 

  - Load Oreoscoptes data by `df <- read.csv("BC_data.csv", stringsAsFactors = FALSE)`. 
  
  - Generate a dataframe `coord` with longitude and latitude in BC_data by 
    
    `coord <- df[,c('decimalLatitude','decimalLongitude')]`.
  
  - Clean the data by removing the observations with NA, `coord <- na.omit(coord)`.
  
  - Visualize the data `plot(coord$decimalLatitude, coord$decimalLongitude)`.
  
  - By using the coordinates function to create a new object with the spatial coordinates and then setting its CRS using proj4string, we can ensure that our spatial analyses are performed correctly, and we can compare our results with other analyses that use the same CRS.
    ```
    coordinates(coord)<- ~decimalLongitude+decimalLatitude
    
    proj4string(coord) <- CRS("+proj=longlat +datum=WGS84")
    
    coord_conv <- spTransform(coord,CRS("+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50
    
                  +lat_2=58.5 +x_0=1000000 +y_0=0 +datum=NAD83 +units=m +no_defs"))
    ```
  
  - Load the covariates data by `load("BC_Covariates.Rda")`.
  
  - Define the observation area, window, by `win = as.owin(DATA$Window)`
  
  - create ppp object with coordination and the window by
  
    ```
    ppp_birds <- ppp(coord_conv@coords[,1], 
    
                 coord_conv@coords[,2], 
                 
                 window=win)
    ```
  
3. Inspecting and Exploring data

  - Check the spread of the bird `plot(ppp_birds)`
  
  - Check homogeneity, perform quardratic test.
    ```
    Q <- quadratcount(bei, nx = 10, ny = 10)
    quadrat.test(Q)
    ```
    
  - *Hot spot analysis* to identify areas of elevated intensity by
    ```
    ## Estimate R
    R <- bw.ppl(ppp_birds)
    
    ## Calculate test statistic
    LR <- scanLRTS(ppp_birds, r = R)
    
    ## Plot the output 
    plot(LR)
    ```  
  - Identify the class of each potential covariate by `sapply(DATA, class)`
  
  - Visualise each potential covariate using methods appropriate to each object class.
    ```
    potential <- c("Window", "Elevation", "Forest", "HFI", "Dist_Water")
    
    for (i in potential){
      plot(DATA[[i]], main = i)
      points(ppp_birds$x, ppp_birds$y, pch = 19, col = "white", cex = 0.6)
      points(ppp_birds$x, ppp_birds$y, pch = 19, col = "black", cex = 0.4)
    }
    ```

  - Check relationships between potential covariates via *kernel estimation* by `rhohat`
    ```
    rho_elev <- rhohat(ppp_birds, DATA$Elevation)
    rho_fore <- rhohat(ppp_birds, DATA$Forest)
    rho_hfi <- rhohat(ppp_birds, DATA$HFI)
    rho_dist_water <- rhohat(ppp_birds, DATA$Dist_Water)
    
    plot(rho_elev, xlim = c(0, max(DATA$Elevation)))
    plot(rho_fore, xlim = c(0, max(DATA$Forest)))
    plot(rho_hfi, xlim = c(0, max(DATA$HFI)))
    plot(rho_dist_water, xlim = c(0, max(DATA$Dist_Water)))
    ```
  
4. Model Building - Regression

  - Scale the covariates by their mean and variance
    ```
    mu_elev <- mean(DATA$Elevation)
    stdev_elev <- sd(DATA$Elevation)
    DATA$Elevation_scaled <- eval.im((Elevation - mu_elev)/stdev_elev, DATA)
    
    mu_water <- mean(DATA$Dist_Water)
    stdev_water <- sd(DATA$Dist_Water)
    DATA$Dist_Water_scaled <- eval.im((Dist_Water - mu_water)/stdev_water, DATA)
    ```
    
  - Establish formula1:
    ```
    formula1 <- ppp_birds ~ Elevation_scaled + I(Elevation_scaled^2) + Dist_Water_scaled + I(Dist_Water_scaled^2)
    ```
    
  - Fit the model with formula1 by `fit1 <- ppm(formula1, data = DATA)`.
  
  - Because the model did not converge, we simplified our formula by excluding *I(Elevation_scaled^2)*.
    ```
    formula2 <- ppp_birds ~ Elevation_scaled + Dist_Water_scaled + I(Dist_Water_scaled^2)
    ```
    
  - Fit the model with formula2 by `fit2 <- ppm(formula2, data = DATA)`.
  
  - Test the statistical significance of coefficients by looking at the Ztest from`fit2` table.
  
  - Identify the formula with coefficients `co <- coef(fit2)`

5. Model Validation

  - Model visualisation
    ```
    plot(fit2,
         se = FALSE,
         superimpose = FALSE)
         
    plot(ppp_birds,
         pch = 16,
         cex = 0.7,
         cols = "white",
         add = TRUE)
    plot(ppp_birds,
         pch = 16,
         cex = 0.5,
         cols = "black",
         add = TRUE)
    ```
    
  - Compute the intensity of a fitted point process model as a function of one of its covariates
    ```
    #Mean disw
    #E_disw <- mean(DATA$Dist_Water_scaled) # just 0

    #Elevational effect on lambda at mean disw
    elev_effect <- effectfun(fit2, "Elevation_scaled", Dist_Water_scaled = 0, se.fit = T)
    
    disw_effect <- effectfun(fit2, "Dist_Water_scaled", Elevation_scaled = 0, se.fit = T)
    
    
    
    #Side by side plotting
    par(mfrow = c(1,2))
    
    #Plot the elevation effect 
    
    plot(elev_effect,
         legend = FALSE,
         main = "Elevation effect at mean distance to water ")
    plot(disw_effect,
         legend = FALSE,
         main = "Distance to Water  effect at mean elevation ")
    ```

  - Quardratic Counting by `quadrat.test(fit2, nx = 2, ny = 4)`
  
  - Partial Residual Plots
    ```
    par_res_elev <- parres(fit2, "Elevation_scaled")
    
    #Calculate the relative intensity as a function of gradient
    par_res_disw <- parres(fit2, "Dist_Water_scaled")
    
    #Side by side plotting
    par(mfrow = c(1,2))
    plot(par_res_elev,
         legend = FALSE,
         lwd = 2,
         main = "",
         xlab = "Elevation_scaled")
    plot(par_res_disw,
         legend = FALSE,
         lwd = 2,
         main = "",
         xlab = "Dist_Water_scaled)")
    ```
  
### Packaged Employed. 

1. sp: A package providing classes and methods for spatial data: points, lines, polygons and grids

2. dplyr: A Grammar of Data Manipulation

3. sf: Create sf object

4. spatstat:  Its main focus is the analysis of spatial patterns of points in two-dimensional space.

5. maptools: Translate and disguise coordinate placing in the real world.

6. rgdal: 

## Results

By looking at Figure 1, we can see that the data is not homogeneous dataset as the birds tend to be clustered in south areas of BC, whereas others have no birds at all, and we conduct a quardratic test to support the inhomogeneous assumption (small p-value). Figure 2 illustrates the hotspot is Osoyoos of the dataset. 

```{r, warning=FALSE, fig.cap= "Oreoscoptes in BC"}
# Check the spread of the bird `ppp_birds`
plot(ppp_birds, main = "Spread of Oreoscoptes")

# Confirm inhomogeneity
Q <- quadratcount(ppp_birds, nx = 10, ny = 10)
quadrat.test(Q)
```

```{r, warning=FALSE, fig.cap= "Hotspot of Oreoscoptes in BC"}
# Hot spot analysis to identify areas of elevated intensity
## Estimate R
R <- bw.ppl(ppp_birds)

## Calculate test statistic
LR <- scanLRTS(ppp_birds, r = R)

## Plot the output 
plot(LR)
```

There are four potential covariates in BC_Covariates.Rda, Elevation, Forest, HFI, and Dist_Water. It's hard to tell the relationship between them and intensity, so we used kernel estimation to detect it. Figure 3 shows that there seems to be two clusters: one group living near the 0 elevation, and another group lives in around 500m, and there is a non-linear relationship between elevation and bird intensity. The narrow bandth in Figure 4 describes a specific preference for habitants in terms of the distance to water, and a non-linear relationship between it and bird intensity as well. We exclude `Forest` and `HFI` here because of wider bandwidths.

```{r, warning=FALSE, fig.cap= "Relationship Diagnosis - Elevation"}
plot(rho_elev, xlim = c(0, max(DATA$Elevation)))
```

```{r, warning=FALSE, fig.cap= "Relationship Diagnosis - Water Distance"}
plot(rho_dist_water, xlim = c(0, max(DATA$Dist_Water)))
```

After deciding to use `Elevation` and `Dist_Water` to be candidate of covariates with quardratic effect, we scaled both variables to prevent disproportional impact on bird intesity to build the regression model with formula1: ppp_birds ~ Elevation_scaled + I(Elevation_scaled^2) + Dist_Water_scaled + I(Dist_Water_scaled^2).

However, when we tried to fit first model, it failed to converge. Hence, we simplified our formula to formula2: ppp_birds ~ Elevation_scaled + Dist_Water_scaled + I(Dist_Water_scaled^2) by excluding the I(Elevation_scaled^2) and built a second model. By the Ztest from the model, we can say that all the coefficients are statistically significant. It suggested that the intensity can be estimated by the following function:

$\lambda(u)$ = exp(`r co[1]` + `r co[2]` * Elevation_scaled + `r co[3]` * Dist_Water_scaled + `r co[4]` * I(Dist_Water_scaled^2))

Figure 5 shows that with these two covariates, the model did not perform well because we can see that there's no bird observations(dots) in high prediction(yellow) area. To see how individual coefficients affect the intensity, please follow figure 6. It indicates that `Elevation` has negative marginal effect before it reaches its mean value (`Elevation_scaled` > 0). On the other hand, `Distance_Water` has greatest marginal effect at 2 standard deviation away from its mean value and then the effect goes down.

```{r, warning=FALSE, fig.cap= "Model Fitting - Fitted Plot"}
plot(fit2,
     se = FALSE,
     superimpose = FALSE)
     
plot(ppp_birds,
     pch = 16,
     cex = 0.7,
     cols = "white",
     add = TRUE)
plot(ppp_birds,
     pch = 16,
     cex = 0.5,
     cols = "black",
     add = TRUE)
```
```{r, warning=FALSE, fig.cap= "Individual Effect"}
#Mean disw
#E_disw <- mean(DATA$Dist_Water_scaled) # just 0

#Elevational effect on lambda at mean disw
elev_effect <- effectfun(fit2, "Elevation_scaled", Dist_Water_scaled = 0, se.fit = T)

disw_effect <- effectfun(fit2, "Dist_Water_scaled", Elevation_scaled = 0, se.fit = T)



#Side by side plotting
par(mfrow = c(1,2))

#Plot the elevation effect 

plot(elev_effect,
     legend = FALSE,
     main = "Elevation Effect")
plot(disw_effect,
     legend = FALSE,
     main = "Distance to Water  Effect")
```
Though by Figure 5 we can see that the model did not fit well to our dataset, we performed a quardratic test to be more sure about it. The small p-value shows that there’s a significant deviation from our model’s predictions. Hence, we used the residual plot to see how to enhance it. 

```{r}
#Run the quadrat test
quadrat.test(fit2, nx = 2, ny = 4)
```
```{r, warning=FALSE, fig.cap= "Residual Plot for Model 2"}
#Calculate the partial residuals as a function of elevation
par_res_elev <- parres(fit, "Elevation_scaled")

#Calculate the relative intensity as a function of gradient
par_res_disw <- parres(fit, "Dist_Water_scaled")

#Side by side plotting
par(mfrow = c(1,2))
plot(par_res_elev,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Elevation_scaled")
plot(par_res_disw,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Dist_Water_scaled)")
```

## Conclusion 
- The partial graph, still does not go well to capture the distribution. 

- The reason is more more the variables selection rather than the model.

- Though `The distance to water` and `elevation` may be explainable in some sense, to accurately fit the location, we may need more co-variate.

- For example, a reasonable guess is the average temperature, as all this species are densely located in the south end of BC. 

- To further improve the fitting, we can incoprate `temperature` to our model 


